EvoSwarm: An Event-based Architecture for Multi-population Optimization Algorithms

### 2020-05-13 draft.tex
Starting Results of Speed

To evaluate the effect of these parameters on the scalability of the system, we
propose  an experiment in which we selected $f_4$ (Skew Rastrigin-Bueche separable) from the
BBOB testbed. This function has been used because it is computationally demanding,
and in higher dimensions has been difficult for PSO \cite{el2009black} and GAs
\cite{nicolau2009application} to solve with the required FEs. As we
are only comparing in terms of execution time, the results from a single function
can be better understood, since there are fewer factors involved.
Following the procedure described in the previous section, we executed 
two sets of experiments with five and ten initial populations,
repeating each experiment 30 times using  1, 2, 4, and 8 workers.  We kept the same 
maximum number of FEs, changing the relevant parameters for this. 
See Tables \ref{tab:params:10} and \ref{tab:params:5} 
for the complete list. 
%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{time_diff}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and populations sizes.  }
  % Which function? - JJ
  \label{fig:spworker}
\end{figure}
%
\begin{table}[]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations}
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrr}
    \hline
              &         & 5 Populations    & 10 Populations \\
     Dimension& Workers & Median of rate & Median of rate  \\
    \hline
          2   & 1w      & 2.96   & 3.15   \\
              & 2w      & 1.84   & 1.72   \\
              & 4w      & 1.22   & \textbf{0.96}   \\
              & 8w      & 1.05   & \textbf{0.59}   \\
    \hline
          3   & 1w      & 6.53   & \textbf{7.76}   \\
              & 2w      & 3.64   & 3.83   \\
              & 4w      & 2.33   & 2.26   \\
              & 8w      & 2.17   &  \textbf{1.36}   \\
    \hline
          5   & 1w      & \textbf{12.33}  & 14.19  \\
              & 2w      & 6.63   & \textbf{7.41}   \\
              & 4w      & 4.25   & 4.34   \\
              & 8w      & 3.67   &  \textbf{2.41}   \\
    \hline
          10  & 1w      & \textbf{31.50}  & 38.30  \\
              & 2w      & \textbf{16.07}  & 19.31  \\
              & 4w      &  9.99   & 10.35  \\
              & 8w      & 8.97   &  \textbf{5.92}  \\
    \hline
          20  & 1w      &  116.35 & 124.51 \\
              & 2w      & \textbf{51.18}  & 63.68  \\
              & 4w      & 32.26  & 33.07  \\
              & 8w      & 28.64  & \textbf{18.25}  \\
    \hline
          40  & 1w      & \textbf{369.89} & 448.89 \\
              & 2w      & \textbf{198.43} & 234.50 \\
              & 4w      & 122.39 &  120.62 \\
              & 8w      & 122.31 & \textbf{63.35} \\
    \hline
  \end{tabular}
\end{table}

@@

In Table~\ref{tab:time} we can see that as expected, and in most cases, 
increasing the number of workers reduces the time
required for the execution of an instance, 
mainly when using the same parameters
(number of populations and number of workers), that is, performance
scales with the number of workers for a fixed problem and number of
sub-populations used. See Table~\ref{tab:speedup:test} for a statistical test 
on speedup, this explain further in the next section.

However, when we compare the two
options, we find some subtle differences: When using one or two
workers, using 5 populations takes less time in most cases. 
In Table~\ref{tab:time} results with p-value of the two-tailed Wilcoxon
signed-ranks test was grater than the level of significance of 0.05 
(i.e., accepting the null hypothesis of inequality) are marked in boldface.
A reason for this could be that the migration of individuals takes more time to
reach the worker. For instance, when a single worker has 10 populations, only
after processing every one of them, it will start receiving mixed populations.  Although, in
all cases, and even a single worker, populations are heterogeneous, this
advantage takes more time to propagate. With four workers, almost the same
results can be achieved regardless of our selection between five or ten
populations. Also, there is not much difference between 4 and 8 workers, when
using five populations, which might indicate that some workers are
idling while populations are being processed by other workers. In
general, we can observe that keeping a ratio close to 1:1, but a bit
better than that, between the number of workers and populations, better exploits the computing
resources available. This also decreases the relation between the
overhead needed to migrate and process incoming and outgoing
population and the total computing time: more populations will need
more time to do that.

% I think this is not the case? 
% Another additional issue is related to the algorithm itself. Many more
% workers than populations will have a wider range of states of
% evolution in the experiment; populations that will be mixed will have
% very different fitness states, which means that a part of them will
% probably have been not so useful, following the intermediate
% disturbance hypothesis. When the number of workers is increased, the
% difference between the state of evolution of populations will be
% intermediate, leading to better results; eventually, when there are as
% many workers as populations, the state will be quite similar, which is
% shown with a smaller gap in the time won adding more workers.

>>>>

To evaluate the effect of these parameters on the scalability of the system, we
propose  an experiment in which we selected $f_4$ (Skew Rastrigin-Bueche separable) from the
BBOB testbed. This function has been used because it is computationally demanding,
and in higher dimensions has been difficult for PSO \cite{el2009black} and GAs
\cite{nicolau2009application} to solve with the required FEs. As we
are only comparing in terms of execution time, the results from a single function
can be better understood, since there are fewer factors involved.
Following the procedure described in the previous section, we executed 
two sets of experiments with five and ten initial populations,
repeating each experiment 30 times using  1, 2, 4, and 8 workers.  We kept the same 
maximum number of FEs, changing the relevant parameters for this. 
See Tables \ref{tab:params:10} and \ref{tab:params:5} 
for the complete list. 
%
\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=\textwidth]{time_wrkr}
  \caption{Boxplot of the median time required for the execution of 30 
  instances of function $f_4$, in seconds, different number of workers and populations sizes.  }
  % Which function? - JJ
  \label{fig:spworker}
\end{figure}
%
\begin{table}[]
  \caption{The median of the time required to finish an instance of the function $f_4$, 30 runs
  with different number of workers and populations}
  \label{tab:time}
  \vspace{0.25cm}
    \centering
    \begin{tabular}{ccrr}
    \hline
              &         & 5 Populations    & 10 Populations \\
     Dimension& Workers & Median of rate & Median of rate  \\
    \hline
          2   & 1w      & \textbf{2.96}   & 3.15   \\
              & 2w      & 1.84   & \textbf{1.72}   \\
              & 4w      & 1.22   & \textbf{0.96}   \\
              & 8w      & 1.05   & \textbf{0.59}   \\
    \hline
          3   & 1w      & \textbf{6.53}   & 7.76   \\
              & 2w      & 3.64   & 3.83   \\
              & 4w      & 2.33   & 2.26   \\
              & 8w      & 2.17   & \textbf{1.36}   \\
    \hline
          5   & 1w      & \textbf{12.33}  & 14.19  \\
              & 2w      & \textbf{6.63}   & 7.41   \\
              & 4w      & 4.25   & 4.34   \\
              & 8w      & 3.67   & \textbf{2.41}   \\
    \hline
          10  & 1w      & \textbf{31.50}  & 38.30  \\
              & 2w      & \textbf{16.07}  & 19.31  \\
              & 4w      & \textbf{9.99}   & 10.35  \\
              & 8w      & 8.97   & \textbf{5.92}   \\
    \hline
          20  & 1w      & \textbf{116.35} & 124.51 \\
              & 2w      & \textbf{51.18}  & 63.68  \\
              & 4w      & 32.26  & 33.07  \\
              & 8w      & 28.64  & \textbf{18.25}  \\
    \hline
          40  & 1w      & \textbf{369.89} & 448.89 \\
              & 2w      & \textbf{198.43} & 234.50 \\
              & 4w      & 122.39 & \textbf{120.62} \\
              & 8w      & 122.31 & \textbf{63.35} \\
    \hline
  \end{tabular}
\end{table}

@@

In Table~\ref{tab:time} we can see that as expected, and in most cases, 
increasing the number of workers reduces the time
required for the execution of an instance (shown in boldface), mainly when using the same parameters
(number of populations and number of workers), that is, performance
scales with the number of workers for a fixed problem and number of
sub-populations used.

However, when we compare the two
options, we find some subtle differences: When using one or two
workers, using 5 populations takes less time in most cases. 
Results 
A 
reason for this could be that the migration of individuals takes more time to
reach the worker. For instance, when a single worker has 10 populations, only
after processing every one of them, it will start receiving mixed populations.  Although, in
all cases, and even a single worker, populations are heterogeneous, this
advantage takes more time to propagate. With four workers, almost the same
results can be achieved regardless of our selection between five or ten
populations. Also, there is not much difference between 4 and 8 workers, when
using five populations, which might indicate that some workers are
idling while populations are being processed by other workers. In
general, we can observe that keeping a ratio close to 1:1, but a bit
better than that, between the number of workers and populations, better exploits the computing
resources available. This also decreases the relation between the
overhead needed to migrate and process incoming and outgoing
population and the total computing time: more populations will need
more time to do that.

Another additional issue is related to the algorithm itself. Many more
workers than populations will have a wider range of states of
evolution in the experiment; populations that will be mixed will have
very different fitness states, which means that a part of them will
probably have been not so useful, following the intermediate
disturbance hypothesis. When the number of workers is increased, the
difference between the state of evolution of populations will be
intermediate, leading to better results; eventually, when there are as
many workers as populations, the state will be quite similar, which is
shown with a smaller gap in the time won adding more workers.
